\ab{Constraints on the quasi-normal mode frequencies of the LIGO-Virgo signals by making full use of gravitational-wave modeling}

### 2021-01-08 papers/intro_paper/intro_paper.tex
changes in Introduction

The LIGO Scientific Collaboration~\citep{lsc} and the Virgo Collaborations~\citep{Virgo} recently announced their catalogue of gravitational wave (GW) observations~\citep{GWTC-2} from the first half of their third observing run (O3a)~\citep{O3reference}. Combined with the confirmed observations from the first and second observing runs~\citep{abbott2019gwtc}, the Advanced LIGO detectors at Hanford, Washington and Livingston, Louisiana~\citep{aasi2015characterization}, and the Advanced Virgo detector in Cascina, Italy~\citep{acernese2014advanced} have now detected more than $50$ GW events from the merger of compact objects like neutron stars and/or black holes (BHs), collectively called compact binary coalescences (CBCs). Alongside independent claims of detections~\citep{nitz20191,nitz20202,2019PhRvD.100b3007Z,2020PhRvD.101h3030V,Venumadhav_2020}, this has firmly established the field of GW astronomy, five years on from the first ever direct detection of GWs on Earth, on September 14, 2015~\citep{abbott2016observation}.
\par
Observation of GWs have had significant astrophysical and cosmological implications~\citep{LSC_2016astroph,gw170817_mma,gw170817_joint,gw170817_hubble}. It has also allowed us to make statements in fundamental physics. Specifically, LIGO-Virgo's observations have allowed us to test predictions of Einstein's theory of General Relativity (GR)~\citep[GR]{}, in previously unexplored regimes of highly relativistic, strong-field regimes of gravity~\citep{LSC_2016grtests,GW170817_TGR,gwtc1_tgr}. In GR, a CBC involving two BHs (a binary black hole (BBH) system) is described in three distinct phases: an early \textit{inspiral}, where the two compact objects spiral in and \textit{plunge} due to a back-reaction of gravitational-wave-emission, a \textit{merger} marked by the formation of an apparent horizon~\citep{NRpaper}, and a late-time \textit{ringdown}, where the newly formed remnant object settles down to a stable Kerr state through the emission of an exponentially damped quasi-normal-mode (QNM) spectrum of gravitational radiation~\citep{vishu,earlyqnmpapers}.  
\par
The LIGO-Virgo collaborations have also released a companion paper detailing their latest results of tests of GR using events from the catalogues GWTC-2 CITE and GWTC-1 CITE. The results include tests of GW generation and source dynamics, where bounds are placed on parametrised deviations in the Post-Newtonian coefficients describing the early inspiral~\citep{earlydevelopmentpapers}, and phenomenological coefficients describing the intermediate (plunge) and merger regimes of coalescence~\citep{TIGERmethodspapers}, tests of GW propagation, which assume a generalised dispersion relation and place upper bounds on the Compton wavelength and consequently, the mass of the graviton~\citep{gw170104,samajdar2017projected} and tests of the polarisation of gravitational radiation using a multi-gravitational-wave-detector network~\citep{gw170814,isi2017probing}. The paper also checks for consistency between different portions of the signal using estimates of predicted mass and spin of the remnant object~\citep{Ghosh:2016xx,Ghosh:2017gfp,LSC_2016grtests}, and consistency of the residuals with interferometric noise~\citep{Ghonge:2020suv,gwtc1_tgr}. None of these tests report any departure from the predictions of GR.
\par
The LIGO-Virgo O3a testing GR paper also details results on tests of BH ringdown and nature of the remnant object, which is an active field of research at the moment. The no-hair conjecture in GR~\citep{} states that an (electrically neutral) astrophysical BH is completely described by two observables: mass and spin angular momentum. One consequence of the no-hair conjecture is that the (complex) QNM frequencies of gravitational radiation emitted by a perturbed isolated BH is uniquely determined by its mass and spin angular momentum. Hence a test of the no-hair conjecture would involve checking for consistency between estimates of mass and spin of the remnant object across multiple QNM frequencies. An inconsistency would either indicate a non-BH nature of the remnant object, or an incompleteness of GR as the underlying theory of gravity. The consistency between the post-merger signal and the least damped QNM was first demonstrated in ~\citep{LSC_2016grtests} for GW150914, and later extended to include overtones in~\citep{Brito:2018rfr,Giesler:2019uxc,Isi:2019aib,Bhagwat:2019dtm,Forteza:2020hbw}. Consistency of the late-time signal with a single QNM is a test of the ringdown of a BBH coalescence, but not necessarily a test of the no-hair conjecture, which requires the measurement of (at least) two QNMs (BH spectroscopy), and checking for consistency between them. Recent work in that direction include~\citep{Carullo:2018gah,Carullo:2019flw,Bhagwat:2019bwv}. The nature of the remnant object has also been explored through tests of BH thermodynamics, like the Hawking's area theorem~\citep{Cabero:2017avf} or through search for echos in the post-merger signal~\citep{Nielsen:2018lkf,Tsang:2019zra,Lo:2018sep,Abedi:2018npz,Abedi:2020sgg,Testa:2018bzd}. None of these tests have found evidence for non-BH nature of the remnant object (as described in GR) in LIGO-Virgo BBH observations.
\par
Most of the tests mentioned above focus on analysing the post-merger or late-time ringdown signal in isolation. Second generation ground-based interferometric detectors like Advanced LIGO and Virgo are most sensitive to stellar-mass BH binaries that merge near the minima of their sensitivity band ($\sim 100$Hz). As a consequence, the remnant object \textit{rings down} in shot-noise dominated higher frequencies, leaving very little signal-to-noise ratio (SNR) in the post-merger signal. The ringdown start time is not a clearly defined quantity and has been explored in detail in~\citep{Bhagwat:2017tkm}. In other works~\citep{Carullo:2018gah,Carullo:2019flw}, it has been left as a free parameter to be estimated directly from the data. Uncertainties in estimates of the ringdown start-time, as well as an overall lack of SNR in the post-merger signal, given typical sensitivities of ground-based detectors result in significant statistical uncertainties in the measurement of the QNM frequencies.
\par
An independent approach to BH spectroscopy, based on a full-signal analysis, was outlined in~\citep{Brito:2018rfr} (henceforth referred to as \paperone). Unlike methods that focus on the post-merger signal, it describes a framework in which a complete inspiral-merger-ringdown (IMR) waveform model is used to measure the complex QNM frequencies. This allows access to the full SNR of the signal, reducing measurement uncertainties. Moreover, the definition of the ringdown start time is built into the model and does not need to be either left as an additional free parameter or fixed using alternate definitions. While \paperone presented the method in the context of a non-spinning waveform model, we extend the analysis to the case with BHs spins in the current paper. \abhi{mention other improvements between EOBNRv2 and EOBNv4} All astrophysical BHs are expected to be spinning, and ignoring effects of spin have been shown to introduce systematic biases in the measurement of the source properties CITE \cite{paper_showing_systematics_from_ignoring_spin}.
\par
The rest of this paper is organized as follows. Section~\ref{sec:model} describes our parametrised IMR waveform model. In section~\ref{sec:method}, we define our testing GR framework in which we use that model to measure the complex QNM frequencies within a Bayesian formalism. Then in section~\ref{sec:results}, we demonstrate our method on real GW events, as well as simulated signals. Finally we provide a summary of our results and discuss future developments in section~\ref{sec:discussion}.

@@

\section*{ACKNOWLEDGEMENTS}\label{sec:acknowledgements}
LIGO Clusters

\appendix
\section{Study of syetematics on ringdown measurements}\label{sec:noise_systematics}

>>>>

\comment{AB: Please make the citations in the bibliography uniform (I would stick with citations 
taken from InSpires and remove the URL links). We will be submitting to Physical Review D, so please 
stick with American English.}

The LIGO Scientific Collaboration~\citep{lsc} and the Virgo
Collaboration\sout{s}~\citep{Virgo} \ab{have} recently announced their catalogue of
gravitational-wave (GW) \sout{observations} \ab{signals}~\citep{GWTC-2} from the first
half of the\sout{ir} third observing run (O3a)~\citep{O3reference}. Combined
with \sout{the confirmed observations from} the first and second observing-run \ab{catalogues}~\citep{abbott2019gwtc}, the Advanced LIGO detectors at Hanford,
Washington and Livingston, Louisiana~\citep{aasi2015characterization},
and the Advanced Virgo detector in Cascina,
Italy~\citep{acernese2014advanced} have now detected \sout{more than} $50$ GW
events from the merger of compact objects like neutron stars and/or
black holes (BHs).\sout{, collectively called compact binary coalescences
(CBCs).} Alongside independent claims of
detections~\citep{nitz20191,nitz20202,2019PhRvD.100b3007Z,2020PhRvD.101h3030V,Venumadhav_2020},
\sout{this} \ab{these results} \sout{has} \ab{have} firmly established the field of GW astronomy, five years \sout{on
from} \ab{after} the first \sout{ever direct} detection of a GW \ab{passing through the} \sout{on} Earth, \ab{notably GW150914} 
\sout{on September 14,2015}~\citep{abbott2016observation}.

\ab{The observation of GWs has} had significant astrophysical and cosmological
implications~\citep{LSC_2016astroph,gw170817_mma,gw170817_joint,gw170817_hubble}. It
has also allowed us to \sout{make statements in} \ab{probe} fundamental
physics \sout{.Specifically, LIGO-Virgo's observations have allowed us to} and 
test predictions of Einstein's theory of General Relativity 
(GR)~\citep{GR} in the previously unexplored \ab{highly-dynamical, and strong field 
regime} \sout{of highly relativistic, strong-field regimes of
gravity}~\citep{LSC_2016grtests,GW170817_TGR,gwtc1_tgr}. In GR\sout{, a CBC
involving two BHs (}a binary black hole (BBH) system\sout{)} is described \sout{in} \ab{by}
three \sout{distinct} phases: an early \textit{inspiral}, where the two
compact objects spiral in \ab{loosing energy because of the emission of GWs} \sout{and \textit{plunge} due to a back-reaction
of gravitational-wave-emission}, a \textit{merger} marked by the
formation of a\sout{n} \ab{common} apparent horizon~\citep{NRpaper}, and a \sout{late-time} 
\textit{ringdown}, \sout{where} \ab{during which} the newly formed remnant object settles down
to a \sout{stable} Kerr \ab{BH} \sout{state through the emission of} \ab{emitting} an exponentially damped
quasi-normal-mode (QNM) spectrum of gravitational
radiation~\citep{vishu,earlyqnmpapers}.

The LIGO-Virgo collaborations \ab{have released companion papers \cite{}} detailing their \sout{latest} results 
of tests of GR using \sout{events} \ab{GW150914 \cite{} and 
the first two transient catalogues (TC) (i.e., GWTC-1\cite{}  and GWTC-2 \cite{})}. 
The results include tests of GW generation and source dynamics, where bounds are placed on
parametrized deviations in the post-Newtonian \ab{(PN)} coefficients describing
the early inspiral~\citep{earlydevelopmentpapers}, and
phenomenological coefficients describing the intermediate (plunge) and
merger regimes of coalescence~\citep{TIGERmethodspapers}; tests of GW
propagation, which assume a generalized dispersion relation and place
upper bounds on the Compton wavelength and consequently, the mass of
the graviton~\citep{gw170104,samajdar2017projected}, and tests of the
polarization of gravitational radiation using a
multi-GW-detector network~\citep{gw170814,isi2017probing}. The \ab{GWTC-1/2} paper\ab{s} also 
check\sout{s} for consistency between different portions of the signal using estimates
\sout{of} \ab{for the} predicted mass and spin of the remnant
object~\citep{Ghosh:2016xx,Ghosh:2017gfp,LSC_2016grtests}, and
consistency of the residuals with interferometric
noise~\citep{Ghonge:2020suv,gwtc1_tgr}. None of these tests report any
departure from the predictions of GR.

\ab{The first paper on tests of GR by the LIGO Collaboration~\cite{} provided us with the 
first measurement of the dominant damped-sinusoidal signal in the ringdown stage of a BBH 
coalescence.} The LIGO-Virgo O3a testing GR paper~\cite{} \ab{has recently reported} 
\sout{also details results on tests} \ab{a comprehensive analysis of the properties of the 
remnant object, including the ringdown stage, for tens of GW events.}   
\sout{BH ringdown and nature of the remnant object, which is an active field
of research at the moment.} The no-hair conjecture in GR~\citep{}
states that an (electrically neutral) astrophysical BH is completely
described by two observables: mass and spin angular momentum. One
consequence of the no-hair conjecture is that the (complex) QNM
frequencies of gravitational radiation emitted by a perturbed isolated
BH is uniquely determined by its mass and spin angular momentum. Hence
a test of the no-hair conjecture would involve checking for
consistency between estimates of mass and spin of the remnant object
across multiple QNM frequencies~\cite{Dreyer:2003bv}. An inconsistency would either
indicate a non-BH nature of the remnant object, or an incompleteness
of GR as the underlying theory of gravity.

The consistency between the post-merger signal and the least damped QNM was first demonstrated in 
Ref.~\citep{LSC_2016grtests} for GW150914, and later extended to include
overtones in Refs.~\citep{Brito:2018rfr,Giesler:2019uxc,Isi:2019aib,Bhagwat:2019dtm,Forteza:2020hbw}. Consistency
of the late-time \sout{signal} \ab{waveform} with a single QNM is a test of the ringdown of
a BBH coalescence, but not necessarily a test of the no-hair
conjecture, which requires the measurement of (at least) two QNMs (BH
spectroscopy), and checking for consistency between them~\cite{Dreyer:2003bv,Berti:2005ys}. Recent work
in that direction \comment{AB: which direction? There are papers by Nikhef group and 
also Brito et al., and many others that have investigated BH spectroscopy.} include~\citep{Carullo:2018gah,Carullo:2019flw,Bhagwat:2019bwv}. The
nature of the remnant object has also been explored through tests of
BH thermodynamics, like the Hawking's area
theorem~\citep{Cabero:2017avf} or through search for echos in the
post-merger
signal~\citep{Nielsen:2018lkf,Tsang:2019zra,Lo:2018sep,Abedi:2018npz,Abedi:2020sgg,Testa:2018bzd}. None
of these tests have found evidence for non-BH nature of the remnant
object (as described in GR) in LIGO-Virgo BBH observations.

\comment{AB: I don't understand the next senetnce, it looks to me out of context.} 
Most of the tests mentioned above focus on analysing the post-merger
or late-time ringdown signal in isolation. \sout{Second generation
ground-based interferometric detectors like} \ab{Ground-based detectors, 
such as \sout{Advanced} LIGO and Virgo,} are most sensitive to 
stellar-mass BH binaries that merge near the
minim\ab{um}\sout{a} of their sensitivity band ($\sim 100$Hz). As a consequence, the
remnant object \textit{rings down} \comment{AB: we are not discussing the interferometer noises, 
so why to give such a detail about the shot noise?} \sout{in shot-noise dominated higher
frequencies, leaving} \ab{at frequencies where the sensitivity is worse, leaving} 
very little signal-to-noise ratio (SNR) in the
post-merger signal. \comment{AB: This discussion about the starting time of the ringdown starts too abruptly, 
and looks out of context and incomplete. Also, why are we citing only those papers? The matter 
has a long history. There have been calibrations to NR of damped-sinusoid signals.} 
The ringdown start time is not a clearly defined
quantity and has been explored in detail
in Ref.~\citep{Bhagwat:2017tkm}. In other
works~\citep{Carullo:2018gah,Carullo:2019flw}, it has been left as a
free parameter to be estimated directly from the data. Uncertainties
in estimates of the ringdown start-time, as well as an overall lack of
SNR in the post-merger signal, given typical sensitivities of
ground-based detectors result in significant statistical uncertainties
in the measurement of the QNM frequencies.

An independent approach to BH spectroscopy, based on the full-signal
analysis, was \sout{outlined} \ab{introduced} in Ref.~\citep{Brito:2018rfr} 
(henceforth referred to as \paperone). Unlike methods that focus \ab{only} 
on the post-merger signal, it \sout{describes a framework in which a} 
\ab{employs the} complete inspiral-merger-ringdown
(IMR) waveform \sout{model is used} to measure the complex QNM
frequencies. This \sout{allows} \ab{gives} access to the full SNR of the signal,
reducing measurement uncertainties. Moreover, the definition of the
ringdown start time is built into the \ab{merger-ringdown} model and does not need to be
either left as an additional free parameter or fixed using alternate
definitions. While \paperone presented the method \ab{and tested it for }\sout{in the context of a
non-spinning waveform model} \ab{non-spinning BBHs}, \ab{here} we extend the analysis to the \ab{more 
realistic astrophysical} case in which BHs \ab{carry} spin\sout{s in the current paper}. \ab{Furthermore, 
the IMR waveform model used in this paper is more accurate than what employed in \paperone, 
because it contains more information from PN theory and it was calibrated to a much larger set 
of numerical-relativity waveforms~\cite{Bohe:2016gbl}.}  All astrophysical BHs are expected to be spinning, 
and ignoring effects of spin have been shown to introduce systematic biases in the measurement of the 
source properties \cite{paper_showing_systematics_from_ignoring_spin}.

The rest of the paper is organized as follows. Section~\ref{sec:model} describes our parametrized IMR
waveform model. In Sec.~\ref{sec:method}, we define our testing GR  framework in which we use that model 
to measure the complex QNM frequencies within a Bayesian formalism. Then, in Sec.~\ref{sec:results}, 
we demonstrate our method on real GW events, as well as simulated signals. \ab{In particular, we analyze 
the ... and constrain the QNM frequecies ...} Finally, we provide a summary of
our results and discuss future developments in Sec.~\ref{sec:discussion}.

@@

\section*{Acknowledgements}
\label{sec:acknowledgements}
LIGO Clusters. \comment{AB: Our cluster!!!}

\appendix

\section{Study of syetematics on ringdown measurements}\label{sec:noise_systematics}
